Toward single particle reconstruction without particle picking: Breaking the detection limit

### 2018-08-28 pnas/Towards.tex
Afternoon session

We compute the average second-order autocorrelation of the micrographs. This is a particularly simple computation which can be efficiently executed with a fast Fourier transform (FFT) on GPUs. Given the noise level and number of image repetitions, the second-order autocorrelation of the image can be easily deduced from~\eqref{eq:ac2_micrograph}.  Then, to estimate the target image, we resort to a standard phase retrieval algorithm called relaxed-reflect-reflect (RRR)~\cite{elser2017rrr}, initialized randomly.

Figure~\ref{fig:Einst_example} shows several estimated images for a growing number of micrographs, and a movie is available in \TODO{supplementary material}. Figure~\ref{fig:error_per_micro} presents the normalized recovery error as a function of the amount of data available. Error is measured as the ratio of the root mean square error (RMSE) to the norm of the ground truth (square root of the sum of squared pixel intensities). This is computed after fixing the reflection symmetries (see Section~\ref{sec:AC_analysis}). As evidenced by these figures, the ground truth image can be estimated increasingly well from increasingly many micrographs, without particle picking.
\TODO{Tamir: the experiment is currently running; should be ready tomorrow  (Aug 28).}

@@

\begin{figure*}[h!]
	\centering
	\includegraphics[scale=0.7]{Einstein_progress_examples}
	\caption{\label{fig:Einst_example} Recovery of Einstein from micrographs at noise level $\sigma = 3$ (see Figure~\ref{fig:micro_example}(c)). Averaged autocorrelations of the micrographs allow to estimate the power spectrum of the target image. This does not require particle picking. A phase retrieval algorithm (RRR) produces the estimates here shown, initialized with an image of the physicist Isaac Newton. Estimates are obtained from $P$ micrographs (growing across panels), each containing $700$ image occurrences on average.
		%\TODO{To add: redo the figures according to Amit's comments and the figures of noisy Einstein's micrographs}
		% At a noise level of $\sigma = 3$, this amounts to an SNR of $1/20$.
	}
\end{figure*}

\begin{figure}[h]
	\centering
	\includegraphics[scale=.55]{Einstein_progress} (left) power spectrum estimation error (right) RRR error after fixed number of 1000 iterations.
	\caption{\label{fig:error_per_micro} Relative root mean square error of the estimate of Einstein's image as a function of the number of observed micrographs (logarithmic scale along both axes.)}
\end{figure}

@@

We also note several aspects of SPR experiments that were neglected in this paper, including contrast transfer function (CTF), non-uniformity of the viewing directions \TODO{We never talk about viewing directions at all, so it's strange to mention that non-uniformity is an issue. I think this is confusing. I'd say something like: "relating the observed autocorrelations to the autocorrelations of the 3-D volume" or something like that.} and other contaminants in the micrographs, such as ice crystals.
 All these aspects must be taken into account so the method could be applied on real data. We hope to take care of these issues in future research, as well to extend the proposed method to conformational heterogeneity. 
%\TODO{another issue is that micrographs also contain contaminants such as carbon, ice crystals, etc.. }

%\TODO{Where and how do we cite Kam? Fred?}

>>>>

We compute the average second-order autocorrelation of the micrographs. This is a particularly simple computation which can be efficiently executed with a fast Fourier transform (FFT) on GPUs. Given the noise level and number of image repetitions, the second-order autocorrelation of the image can be easily deduced from~\eqref{eq:ac2_micrograph}.  Then, to estimate the target image, we resort to a standard phase retrieval algorithm called relaxed-reflect-reflect (RRR)~\cite{elser2017rrr}, initialized randomly.
Relative error is measured as the ratio of the root mean square error to the norm of the ground truth (square root of the sum of squared pixel intensities).

Figure~\ref{fig:Einst_example} shows several estimated images for a growing number of micrographs, and a movie is available in \TODO{supplementary material}. Figure~\ref{fig:error_per_micro} presents the normalized recovery error as a function of the amount of data available.  This is computed after fixing the reflection symmetries (see Section~\ref{sec:AC_analysis}). As evidenced by these figures, the ground truth image can be estimated increasingly well from increasingly many micrographs, without particle picking.
\TODO{Tamir: the experiment is currently running; should be ready tomorrow  (Aug 28).}

@@

\begin{figure*}[h]
	\centering
	\begin{subfigure}[h]{0.24\textwidth}
		\centering
		\includegraphics[scale=0.35]{Einstien_progress_1}
		\caption{$5\times 10^2$ micrographs}
	\end{subfigure} 
	\begin{subfigure}[h]{0.24\textwidth}
	\centering
	\includegraphics[scale=0.35]{Einstien_progress_10}
	\caption{$5\times 10^3$ micrographs}
\end{subfigure} 
	\begin{subfigure}[h]{0.24\textwidth}
		\centering
		\includegraphics[scale=0.35]{Einstien_progress_100}
		\caption{$5\times 10^4$ micrographs}
	\end{subfigure} 
	\begin{subfigure}[h]{0.24\textwidth}
		\centering
		\includegraphics[scale=0.35]{Einstien_progress_246}
		\caption{$500\times 246$ micrographs}
	\end{subfigure}
	\caption{\label{fig:Einst_example} Recovery of Einstein from micrographs at noise level $\sigma = 3$ (see Figure~\ref{fig:micro_example}(c)). Averaged autocorrelations of the micrographs allow to estimate the power spectrum of the target image. This does not require particle picking. A phase retrieval algorithm (RRR) produces the estimates here shown, initialized randomly. Estimates are obtained from $5\times 10^2,5\times 10^3,5\times 10^4$ micrographs (growing across panels), each containing $700$ image occurrences on average. \TODO{XP is still running}}	
\end{figure*}

%\begin{figure*}[h!]
%	\centering
%	\includegraphics[scale=0.7]{Einstein_progress_examples}
%	\caption{\label{fig:Einst_example} Recovery of Einstein from micrographs at noise level $\sigma = 3$ (see Figure~\ref{fig:micro_example}(c)). Averaged autocorrelations of the micrographs allow to estimate the power spectrum of the target image. This does not require particle picking. A phase retrieval algorithm (RRR) produces the estimates here shown, initialized with an image of the physicist Isaac Newton. Estimates are obtained from $P$ micrographs (growing across panels), each containing $700$ image occurrences on average.
%		%\TODO{To add: redo the figures according to Amit's comments and the figures of noisy Einstein's micrographs}
%		% At a noise level of $\sigma = 3$, this amounts to an SNR of $1/20$.
%	}
%\end{figure*}

\begin{figure}[h]
	\centering
\begin{subfigure}[h]{0.23\textwidth}
	\centering
	\includegraphics[scale=0.33]{Einstein_ps_error}
	\caption{relative power spectrum estimation error}
\end{subfigure} 
\begin{subfigure}[h]{0.23\textwidth}
	\centering
	\includegraphics[scale=0.33]{Einstein_recovery_error}
	\caption{relative recovery error}
\end{subfigure} 
	\caption{\label{fig:error_per_micro} Error curves for the experiment of Figure~\ref{fig:Einst_example} \TODO{XP is still running}}
\end{figure}

%
%\begin{figure}[h]
%	\centering
%	\includegraphics[scale=.55]{Einstein_progress} (left) power spectrum estimation error (right) RRR error after fixed number of 1000 iterations.
%	\caption{\label{fig:error_per_micro} Relative root mean square error of the estimate of Einstein's image as a function of the number of observed micrographs (logarithmic scale along both axes.)}
%\end{figure}

@@

We also note several aspects of SPR experiments that were neglected in this paper, including contrast transfer function (CTF), non-uniformity of the viewing directions \TODO{We never talk about viewing directions at all, so it's strange to mention that non-uniformity is an issue. I think this is confusing. I'd say something like: "relating the observed autocorrelations to the autocorrelations of the 3-D volume" or something like that.} and other contaminants in the micrographs, such as ice crystals.
 All these aspects must be taken into account so the method could be applied on real data. We hope to take care of these issues in future research, as well to extend the proposed method to conformational heterogeneity.
